\documentclass[11pt]{article}

% --- arxiv-style template (kourgeorge/arxiv-style) ---
\usepackage{arxiv}

% --- engine: xelatex (Unicode-friendly) ---
\usepackage{fontspec}
\setmainfont{Times New Roman}
\setmonofont{Menlo}[Scale=0.85]

% --- core packages ---
\usepackage{amsmath, amssymb}
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{xcolor}
\usepackage{hyperref}
\hypersetup{
    colorlinks=true,
    linkcolor=blue,
    citecolor=blue,
    urlcolor=blue,
}
\usepackage{graphicx}
\usepackage{microtype}
\usepackage{caption}

% --- url-aware line breaking ---
\usepackage{url}
\PassOptionsToPackage{hyphens}{url}
\Urlmuskip=0mu plus 1mu      % allow stretch around URL break-points
\sloppy                       % prefer loose lines over overflow
\setlength{\emergencystretch}{3em}

% --- inline-code (pandoc \passthrough) breakability ---
% Long file paths and identifiers that pandoc renders as
% \texttt{...} are otherwise unbreakable, which produced 200+pt
% overfull \hbox warnings. \seqsplit inserts a discretionary
% break opportunity between every character; LaTeX only breaks
% there when no better option exists, so short inline-code spans
% stay intact in the normal flow.
\usepackage{seqsplit}

% --- code listings ---
\usepackage{listings}
\lstset{
    basicstyle=\ttfamily\small,
    breaklines=true,
    columns=fullflexible,
    keepspaces=true,
}

% --- bibliography (bibtex, more portable than biber) ---
\usepackage[numbers]{natbib}

% --- pandoc compatibility helpers ---
\usepackage{calc}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\providecommand{\pandocbounded}[1]{#1}
\providecommand{\real}[1]{#1}
% pandoc emits \texttt{...} (and sometimes \passthrough{...}) around
% inline-code spans. Override BOTH so long file paths and identifiers
% break across lines instead of producing 200+pt overfull \hbox
% warnings. \seqsplit only inserts break opportunities; LaTeX only
% breaks there when no better option exists, so short inline-code
% spans stay intact in the normal flow.
\newcommand{\passthrough}[1]{\texttt{\seqsplit{#1}}}
\let\OldTexttt\texttt
\renewcommand{\texttt}[1]{\OldTexttt{\seqsplit{#1}}}

% Long tables (Tables 11/12 with 7-column engine grids carry rows
% like "707 302 µs" + "not reported by runner" + "39.06 MB †" that
% otherwise overflow column widths). Render every longtable in
% \footnotesize so the wide rows fit; the surrounding prose stays
% in 11pt body size.
\usepackage{etoolbox}
\AtBeginEnvironment{longtable}{%
  \scriptsize%                   % was \footnotesize — clears the
                                  % last 1pt overflow in Table 12
  \setlength{\tabcolsep}{3pt}% tighter cell padding (default 6pt)
  \renewcommand{\arraystretch}{0.95}% tighter row spacing
}
% pandoc emits \def\LTcaptype{none} around longtables to suppress
% caption numbering; longtable then tries \addtocounter{\LTcaptype}.
% Define the dummy counter so tectonic does not bail.
\newcounter{none}

% --- title block ---
\title{Dazzle: An Embedded Database for On-Device LLM Agents}

\author{
    Ivan Aliaga \\
    Universidad Ricardo Palma \\
    Lima, Peru \\
    \texttt{ivan.aliaga@urp.edu.pe}
}

\date{April 2026}

% --- arxiv metadata ---
\renewcommand{\headeright}{Preprint}
\renewcommand{\undertitle}{Preprint}
\renewcommand{\shorttitle}{Dazzle: An Embedded Database for On-Device LLM Agents}

% --- begin document ---
\begin{document}

\maketitle

\begin{abstract}
Mobile applications that embed a language-model agent now need a
persistent state layer between inference calls --- sensor windows,
materialised aggregates, approximate counts, vector indices over
recent context. The embedded databases available on the platform
(SQLite, RocksDB, LMDB, ObjectBox) ship two to four data-structure
primitives each, so every higher-order construct (bounded streams,
per-field TTL, HyperLogLog, HNSW vector search) is reimplemented
in application code, on top of those primitives, by every team
that needs them.

We present \textbf{Dazzle}, a fork of Valkey (the Linux
Foundation's Redis-compatible store) whose server runs
\emph{inside} the process of Android and iOS applications rather
than as a daemon over TCP loopback. The fork ships ten native
data-structure primitives, including HNSW vector search, against
$\leq 4$ in any embedded competitor we measure on the same
platforms; it exposes them through a typed mobile SDK on Maven
Central, pub.dev, npm and Swift Package Manager.

We characterise the latency envelope across twelve backends on
two physical devices (Moto G35 5G, iPhone 12 Pro). Steady-state
retrieval on the materialised-aggregates path lands at 50~µs on
Moto G35 5G and 7.24~µs on iPhone 12 Pro; HNSW vector search
stays sub-millisecond at recall $\geq 0.95$ across the
9-configuration grid we measure. At 50~µs against a
$\sim 2.84$~s on-device LLM inference turn the retrieval cost is
\textbf{0.00176\,\%} of the inference call --- below the noise
floor of the end-to-end loop. The deciding axis for an on-device
agent backend is therefore \textbf{primitive surface and
developer ergonomics}, not microbenchmark deltas. We report the
absolute per-engine microsecond numbers anyway, because the same
engines also serve non-LLM workloads (analytics, sync, indexing)
where microseconds matter.

An end-to-end 2$\times$2 ablation on 200 Natural Questions queries
on the Moto G35 5G shows that the retrieval primitive is what
drives factual accuracy at this scale. Qwen 2.5 0.5B with Dazzle
RAG reaches 0.630 \texttt{EM\_contains} against 0.105 without
retrieval ($6.0\times$); Qwen 2.5 1.5B with RAG reaches 0.735
against 0.110 without ($6.7\times$). A 380~MB model with
retrieval beats a 940~MB model ($3\times$ larger) without
retrieval on every factual metric, end-to-end on a
\$150-class Android phone. The bottleneck is not the model size,
the SoC, or the microsecond cost of retrieval; it is whether the
backend lets the agent loop reach a vector index at all.
\end{abstract}

\keywords{embedded database \and on-device LLM \and mobile agents
  \and Valkey \and HNSW vector search \and Apache 2.0}

% --- body content (generated from paper_v2_en.md by pandoc) ---
\input{body.tex}

% --- bibliography ---
\bibliographystyle{unsrt}
\bibliography{refs}

\end{document}
